---
title: "Patterns of climate change in the USA: an analysis of temperature and precipitation trends since 1950"
author: Emily Arsenault, Department of Ecology & Evolutionary Biology, University of Kansas, Lawrence, Kansas 66047 USA
output: pdf_document
fontsize: 12pt
bibliography: datacarpentry.bib
csl: science.csl
---
<!--Chunk for reproducibility checkpoint-->
```{r checkpoint, echo=F, include=F}
#install.packages("checkpoint")
library(checkpoint)
checkpoint("2018-03-01")
```

<!--Chunk for importing the data-->
```{r import, include=F, cache=T}

##### MODULARITY 8: RECORDING DEPENDENCIES AND OUTPUTS FOR EACH CHUNK #####
## Dependencies of chunk: external script "ACISimport.R" and/or files "USAAnnualTemp1950_2016.rds" and "USAAnnualPcpn1950_2016.rds"

##### MODULARITY 7: WRITE PSEUDOCODE #####
#if the file exists (TRUE)
  #read it in
#if the file does not exists (FALSE)
  #use func in external script to download data from the web and create this file

##### MODULARITY 9: USING EXTERNAL SCRIPT #####
source("ACISimport.R") #function for pulling climate data from web

##### MODULARITY 2: WRITING EXTENSIBLE CODE #####
data.temp <- "USAAnnualTemp1950_2008.rds"
data.pcpn <- "USAAnnualPcpn1950_2008.rds"
data.temp.web <- "USAannualTemp1950_2016.rds"

acis.pcpn <- readRDS(data.pcpn)

provided <- FALSE
if(provided) {
  acis.temp <- readRDS(data.temp)
} else if(file.exists(data.temp.web)) {
  acis.temp <- readRDS(data.temp.web)
} else {
  acis.temp <- acis2df(sdate = "19500101", edate = "20161231", states = state.abb)
}

#test
min(acis.temp$year) #start year, 1950
max(acis.temp$year) #end year, 2016
min(acis.pcpn$year) #start year, 1950
max(acis.pcpn$year) #end year, 2008

acis.temp <- na.omit(acis.temp)
acis.pcpn <- na.omit(acis.pcpn)

#Outputs: variable for rds data files (acis.temp; acis.pcpn) imported, loaded into R, and saved to directory as variables
```

<!--Chunk for cleaning the data-->
```{r cleaning, include=F}

#Dependencies: imported dataset (temp)

#Pseudocode
#set variable for minimum number of years needed to include weather station
#use aggregate function to find length (years) for each weather station
#if

#minyears <- 40 #minimum number of years needed to use weather station in subsequent analyses
#acis.temp.stn <- aggregate(year ~ lat + lon + state, data = acis.temp, FUN = length)
#acis.temp.years <- as.data.frame(acis.temp$year[acis.temp.stn$year>minyears]) #only include weather stations that have measured more than x years of data (minyears)
#merge(acis.temp.stn, acis.temp.years)

#acis.temp <- merge(acis.temp.stn, acis.temp.years, by = c(lat, lon, state))

######
#change all year < 40 to NA, do an no.omit
#####

#Outputs: cleaned up dataset (tempclean)

```

<!--Chunk for data analysis-->
```{r analysis, include=F}

#Dependencies: imported, cleaned dataset (tempclean)

##### MODULARITY 1:GIVING VARIABLE NAMES TO CONSTANTS #####
sdate <- 1950
edate.pcpn <- 2008
edate.temp <- 2016

#use apply function to calculate mean temp for each year

#Outputs: mintemp1950, mintemp2016, maxtemp1950, maxtemp2016, maxtemp1950sp, maxtemp2016sp, meantemp1950, meantemp2016, side effect plot of temperature averages by year
```

<!--Chunk for making the map-->
```{r map, include=F}
```

## Abstract

- Applied Climate Information System database (ACIS)
- parameters included :

## Introduction

Periods of warming and cooling are natural phenomena throughout the history of the earth. However, evidence suggests that the unprecedented climate changes occuring over the last century have been a result of anthropogenic increases in carbon dioxide to the atmopshere. Here, we provide evidence to support climate warming in the United States based on patterns of temperature and precipitation. The objectives of this study were to (1) determine whether or not the United States has expereinced climate warming since 1950, (2) highlight key areas of the United States that have exeperienced the greatest changes in average annual temperature, and (3) determine the degree to which precipitation has changed in the United States over the last century.

## Methods

Temperature and precipitation datasets were pulled from the Applied Climate Information System (ACIS) database @Hubbard2004. The temperature dataset includes yearly average temperature data gathered from weather stations located all over the USA, between the years of `r min(acis.temp$year)` and `r max(acis.temp$year)`. The precipitation dataset includes annual average rainfall measurements at weather stations all across the USA, measured between the years of __ and __.

## Results

- figure  
- map

## Discussion

## References
