---
title: "Patterns of climate change in the USA: an analysis of temperature and precipitation trends since 1950"
author: Emily Arsenault, Department of Ecology & Evolutionary Biology, University of Kansas, Lawrence, Kansas 66047 USA
output: pdf_document
fontsize: 12pt
bibliography: datacarpentry.bib
csl: science.csl
---
<!--Chunk for package reproducibility-->
```{r checkpoint, echo=F, include=F}
#install.packages("checkpoint")
#install.packages("plyr") #working with datasets
#install.packages("mapdata") #mapping
#install.packages("RColorBrewer") #color spectra
#install.packages("maps") #mapping
#install.packages("httr") #get patch post head put delete
#install.packages("lubridate") #working with dates
#install.packages("qdapRegex") #regular expression tools
#install.packages("jsonlite") #working with json data

library(reshape2)
library(maps)
library(mapdata)
library(RColorBrewer)
library(classInt)
library(checkpoint)
checkpoint("2018-03-01")
```

<!--Chunk for importing the data-->
```{r import, include=F, echo=F, cache=T}

##### MODULARITY 8: RECORDING DEPENDENCIES AND OUTPUTS FOR EACH CHUNK #####

#Dependencies: external script "ACISimport.R" and/or files "USAAnnualTemp1950_2016.rds" and "USAAnnualPcpn1950_2016.rds"

##### MODULARITY 7: WRITING PSEUDOCODE #####
#source acis2df function from external script
#flag (TRUE/FALSE)
#if flag is TRUE, read in rds file
  #if FALSE, then we want to use a better file
    #if it already exists in our directory, read it in
    #else, use function acis2df to pull it directly from the web

##### MODULARITY 9: USING EXTERNAL SCRIPT #####
source("acis2df.R") #function for pulling ACIS data from web

##### MODULARITY 2: WRITING EXTENSIBLE CODE #####
provided <- FALSE
if(provided) {
  acis.temp <- readRDS("USAAnnualTemp1950_2008.rds")
} else if(file.exists("USAAnnualTemp1950_2016.rds")) {
  acis.temp <- readRDS("USAAnnualTemp1950_2016.rds")
} else {
  acis.temp1 <- acis2df(sdate = "19500101", edate = "19821231", states = state.abb) #first half of the data
  acis.temp2 <- acis2df(sdate = "19830101", edate = "20161231", states = state.abb) #second half of the data
  acis.temp <- cbind(acis.temp1, acis.temp2)
}

acis.pcpn <- readRDS("USAAnnualPcpn1950_2016.rds")

#Outputs: variable for rds data files (acis.temp; acis.pcpn) imported, loaded into R, and saved to directory as variables
```

<!--Chunk for removing weather stations with too few years of data-->
```{r cleaning, include=F}

#Dependencies: imported dataset (temp)

##### MODULARITY 7: WRITING PSEUDOCODE #####
#set variable (minyears) for minimum number of years needed to include weather station
#use aggregate function to find length (years) for each weather station
#if the number of rows for the weather station is less than minyears, give NA value to year
#create a new data frame that omits the weather stations with less than minyears data
#na.omit

##### MODULARITY 1: GIVING VARIABLE NAMES TO CONSTANTS #####
minyears <- 40 #minimum number of years needed to use weather station in subsequent analyses
acis.temp.stn <- aggregate(year ~ variable + lat + lon + state, data = na.omit(acis.temp), FUN = length)
acis.temp.stn$variable[acis.temp.stn$year<minyears] <- NA #give NA values to values we are throwing out
newdf <- na.omit(data.frame(year=acis.temp.stn$year, variable=rep(acis.temp.stn$variable, sapply(acis.temp.stn$variable, length)))) #make new dataframe, omits weather stations with too few years of data
stations.temp <- newdf$variable #all of the variable names that we want
clean.temp <- na.omit(acis.temp)[is.element(na.omit(acis.temp)$variable, stations.temp),] #to use for subsequent analyses

acis.pcpn.stn <- aggregate(year ~ name + lat + lon + state, data = na.omit(acis.pcpn), FUN = length)
acis.pcpn.stn$name[acis.pcpn.stn$year<minyears] <- NA #give NA values to values we are throwing out
newdf <- na.omit(data.frame(year=acis.pcpn.stn$year, name=rep(acis.pcpn.stn$name, sapply(acis.pcpn.stn$name, length)))) #make new dataframe, omits weather stations with too few years of data
stations.pcpn <- newdf$name #all of the variable names that we want
clean.pcpn <- na.omit(acis.pcpn)[is.element(na.omit(acis.pcpn)$name, stations.pcpn),] #to use for subsequent analyses

#I have copied and pasted temp script for pcpn - this could have been made more modular by writing a function to repeat the work for me

#Outputs: cleaned up dataset (clean.temp, clean.pcpn)
```

<!--Chunk for changing format from data frame to matrix-->
```{r matrix, include=F}

#Dependencies: cleaned datasets (newstationtemp, newstationpcpn)

matrix.temp <- na.omit(acast(clean.temp, variable~year, value.var="value"))
dim(matrix.temp) #71, 67

matrix.pcpn <- na.omit(acast(clean.pcpn, name~year, value.var = "data"))
dim(matrix.pcpn) #110, 67

#Outputs: data in matrix form (matrix.temp, matrix.pcpn)
```

## Abstract

An analysis of United States weather station climate data collected between `r min(acis.temp$year)` and `r max(acis.temp$year)` suggests that anthropogenic climate change has had an effect on average annual temperature and precipitation across the United States.


## Introduction

Periods of warming and cooling are natural phenomena throughout the history of the earth. However, evidence suggests that the unprecedented climate changes occuring over the last century have been a result of anthropogenic increases in carbon dioxide to the atmopshere. Here, we provide evidence to support climate warming in the United States based on patterns of temperature and precipitation. The objectives of this study were to (1) determine whether or not the United States has expereinced climate warming since 1950, (2) highlight key areas of the United States that have exeperienced the greatest changes in average annual temperature, and (3) determine the degree to which precipitation has changed in the United States over the last century.

## Methods

Average annual temperature and precipitation data from weather stations across the United States were reported by the American Meteorological Society through the Applied Climate Information System (ACIS) @Hubbard2004. Temperature and precipitation data was collected between the years of `r min(acis.temp$year)` and `r max(acis.temp$year)`. Weather stations with less than `r minyears` total years of data were not considered as part of subsequent analyses. In total, `r length(unique(acis.temp$year))` years of climate data (`r min(acis.temp$year)`-`r max(acis.temp$year)`) from `r nrow(matrix.temp)` weather stations were included as part of the analysis.
All statistical analyses were conducted in R @R2014.

## Results

<!--Chunk for data analysis-->
```{r analysis, echo=F, warning=F}
#Dependencies: matrix converted from dataframe (matrix.temp, matrix.pcpn)

##### MODULARITY 1:GIVING VARIABLE NAMES TO CONSTANTS #####
avg.temp <- (colMeans(matrix.temp))
avg.pcpn <- (colMeans(matrix.pcpn))
year <- 1950:2016

temp.lm <- lm(avg.temp~year, data = as.data.frame(matrix.temp))
temp.sum <- summary(temp.lm)

pcpn.lm <- lm(avg.pcpn~year, data = as.data.frame(matrix.pcpn))
pcpn.sum <- summary(pcpn.lm)

#Outputs: linear regression statistics for temp and pcpn, (side effect) scatter plots for temp and pcpn, variabl
```

Between `r min(acis.temp$year)` and `r max(acis.temp$year)`, mean annual temperature ranged from `r round(min(avg.temp), digits = 2)` to `r round(max(avg.temp), digits = 2)` &deg;F, averaging `r round(mean(matrix.temp), digits = 2)` &deg;F, while average annual precipitation ranged from `r round(min(avg.pcpn), digits = 2)` to `r round(max(avg.pcpn), digits = 2)` inches with a mean level of `r round(mean(matrix.pcpn), digits = 2)` inches overall. A significant, positive relationship was observed between year and average temperature (F = `r round(temp.sum$fstatistic, digits = 2)`, p<0.001, R^2=0.40). A significant relationship was also observed between year and average precipitation (F = `r round(pcpn.sum$fstatistic, digits = 2)`, p<0.001, R^2=0.04).

<!--Chunk for plots-->
```{r plots, echo=F, warning=F}

#Dependencies: avg.
plot(x = year, y = avg.temp, type = "b", col = "red", 
     xlab = "", ylab = "Average Annual Temperature (&deg;F)", xaxt = "n")
plot(x = year, y = avg.pcpn, type = "b", col = "blue",
     xlab = "Year", ylab = "Average Annual Precipitation (in)")

```

<!--Chunk for making 1950 map-->
```{r 1950, echo=F, warning=F}

#Dependencies: acis.temp and acis.pcpn dataframes

sdate <- 1950
map(database = "state")
plotvar <- acis.temp$value[acis.temp$year==sdate]
nclr <- 8
plotclr <- brewer.pal(nclr,"PuOr")
plotclr <- plotclr[nclr:1] # reorder colors
class <- classIntervals(plotvar, nclr, style="equal")
colcode <- findColours(class, plotclr)
points(acis.temp$lon[acis.temp$year==sdate], acis.temp$lat[acis.temp$year==sdate], pch=20, col=colcode, cex=1)

#Output: Tried making a map but wish I could have figured out how to make 1 map with 2016-1950 temperature differences at each weather station
```

<!--Chunk for making 2016 map-->
```{r 2016, echo=F, warning=F, fig.cap="Relative annual average temperature in the United States in 1950 (top) and 2016 (bottom)"}

#Dependencies: acis.temp and acis.pcpn dataframes

edate <- 2016
map(database = "state")
plotvar <- acis.temp$value[acis.temp$year==edate]
nclr <- 8
plotclr <- brewer.pal(nclr,"PuOr")
plotclr <- plotclr[nclr:1] # reorder colors
class <- classIntervals(plotvar, nclr, style="equal")
colcode <- findColours(class, plotclr)
points(acis.temp$lon[acis.temp$year==edate], acis.temp$lat[acis.temp$year==edate], pch=20, col=colcode, cex=1)

#Output: Tried making a map but wish I could have figured out how to make 1 map with 2016-1950 temperature differences at each weather station. I also cannot figure out how to get the maps to line up in the knitted pdf
```

## Discussion

The results of this study show significant changes in average annual temperature and precipitation in the United States between the years of `r min(acis.temp$year)` and `r max(acis.temp$year)`, which provides evidence of the widespread effects of climate change previously reported @Loarie2009. Increases in temperature is expected to have drastic effects on our natural communities and economy @Schlenker2009. While average annual precipitation was relatively similar between year 1950 and year 2016, we expect that precipitation changes are seen at a smaller scale, with less predictable, light rainfall and increasing variability in the form of alternating droughts and floods @Trenberth2003.

## References
