---
title: "Patterns of climate change in the USA: an analysis of temperature and precipitation trends since 1950"
author: Emily Arsenault, Department of Ecology & Evolutionary Biology, University of Kansas, Lawrence, Kansas 66047 USA
output: pdf_document
fontsize: 12pt
bibliography: datacarpentry.bib
csl: science.csl
---
<!--Chunk for reproducibility checkpoint-->
```{r checkpoint, echo=F, include=F}
#install.packages("checkpoint")
library(checkpoint)
checkpoint("2018-03-01")
```

<!--Chunk for loading necessary packages-->
```{r packages, include=F}
#install.packages("maps") #creating maps
#install.packages("httr") #get patch post head put delete
#install.packages("lubridate") #working with dates
#install.packages("qdapRegex") #regular expression tools
#install.packages("jsonlite") #working with json data
library(maps)
library(httr)
library(lubridate)
library(qdapRegex)
library(jsonlite)
```

<!--Chunk for importing the data-->
```{r import, include=F, cache=T}

#Dependencies: if TRUE, rds files / if FALSE, pull directly from acis website
#Outputs: data rds files (temp; pcpn)

#PSEUDOCODE DATA IMPORT
#flag <- T/F
#if flag is true
#import rds data file
#else
#import data from ACIS website

#does this file exist?
#if yes, read it in
#if no, use this function to download it from the web

source("ACISimport.R") #modularity 

provided <- exists("USAAnnualTemp1950_2016.rds") #modularity 2: writing extensible code
if(provided){
  temp <- readRDS("USAAnnualTemp1950_2016.rds")
} else {
  temp <- acis2df(sdate = "19500101", edate = "20161231", states = state.abb)
}
```

<!--Chunk for cleaning the data-->
```{r cleaning, include=T}
temp
```

<!--Chunk for making the map-->
```{r map, include=F}
```

## Abstract

- Applied Climate Information System database (ACIS)
- parameters included :

## Introduction

- Can we see climate warming for ourselves by analyzing this data?  
- In what parts of the country are temperatures getting warmer? Are there any parts that actually got colder over 1950-2010?  
- Can we say, based on these data, what change has occurred in precipitation over the period?

## Methods

## Results

- figure  
- map

## Discussion

## References
