---
title: "Trophic relationships in a Mississippi River macrosystem"
author: "Emily Arsenault"
date: "April 9, 2018"
output: pdf_document
---
<!--Chunk for package reproducibility-->
```{r checkpoint, echo=F, include=F}

#install.packages("checkpoint")
#install.packages("plyr") #working with datasets

#library(classInt)
#library(checkpoint)
#checkpoint("2018-03-01")

```

<!--Chunk for importing the data-->
```{r import, echo=F}

read.csv("muss.csv", header = T)

```

<!--Chunk for cleaning up the data-->
remove NA vals from isotope data, but need those data for richness/abundance still (or not, since richness was not accurately sampled anyways)
```{r, echo=F}

```

## Abstract

## Introduction

Looking at an old data set from the Mississippi River, POC study but never analyzed

They determined several different hydrogeomorpholoical zones (FPZs) of the Missisippi River and collected mussels and fish from each of these sites

Want to understand effects of FPZ on species richness, food chain length, food web structure (Layman's metrics)
How does food chain length (FCL) vary with ecosystem size?
  Calculate baseline primary producer d15N signature
  Use that to calculate trophic positions for each individual fish
  Use that to find maximum trophic position (FCL) at each site
  
## Methods

## Results

## Discussion

## Literature Cited