---
title: "Tests"
author: "Emily Arsenault"
date: "March 28, 2018"
output: pdf_document
---

<!--***DAN: Search for ***DAN to find comments-->

<!--***DAN: This knitted right away, so well done on that.-->


```{r, echo=T}
source("base_sig.R")
source("tp_calc.R")
source("fcl_calc.R")
source("plot_web.R")
```

## Tests for base_sig

Test that it can find the correct primary producer signature mean for site "WIN"

```{r base_sig_test_1, echo=T}
sitetest <- "WIN"
frac <- 3.4
h <- base_sig(muss, frac)
if(round(h[sitetest,2], digits = 3)==7.287)
{
  print("passed")
} else
{
  print("failed")
}
```

## Tests for tp_calc

Test to make sure that no fish trophic position is less than the minimum trophic position that we observed or greater than the maximum trophic position observed (if it is, it was calculated incorrectly)

```{r tp_calc_test_1, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(h$tp>1.841 && h$tp<4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

Tests that it calculates tp correctly for fish id 136

```{r tp_calc_test_2, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(round(h$tp[h$id==136], digits=3)==3.166)
{
  print("passed")
} else
{
  print("failed")
}
```
  
## Tests for fcl_calc

Test to make sure that FCL was calculated correctly for site "SCR"

```{r fcl_calc_test_1, echo=T}
tpdf <- na.omit(tp_calc(fish, 3.4))
h <- fcl_calc(tpdf)
if(round(h$fcl[h$site=="SCR"], digits=3)==4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

## I wonder if there is a way that I can write a unit test for plot_web(), my isotope biplot? Still thinking about this.