---
title: "Tests"
author: "Emily Arsenault"
date: "March 28, 2018"
output: pdf_document
---

<!--***DAN: Search for ***DAN to find comments - all files can have comments-->

<!--***DAN: This knitted right away, so well done on that.-->

<!--***DAN: Great use of git, inc tags-->

<!--***DAN: Your unit tests look great in the sense they are set up as requested, and you have function specs but incomplete function code at the halfway tag, so great job on that! I hoped you learned the benefits! You function specs were pretty good but not perfect. See comments in the .R files. Basically, descriptions of inputs were not unambiguously interpretable in all cases.-->

<!--***DAN: Overall great job! Forge ahead with the final project, and I hope the analysis turns out to be scientifically useful! 10/10-->


```{r, echo=T}
source("base_sig.R")
source("tp_calc.R")
source("fcl_calc.R")
source("plot_web.R")
```

## Tests for base_sig

Test that it can find the correct primary producer signature mean for site "WIN"

```{r base_sig_test_1, echo=T}
sitetest <- "WIN"
frac <- 3.4
h <- base_sig(muss, frac)
if(round(h[sitetest,2], digits = 3)==7.287)
{
  print("passed")
} else
{
  print("failed")
}
```

## Tests for tp_calc

Test to make sure that no fish trophic position is less than the minimum trophic position that we observed or greater than the maximum trophic position observed (if it is, it was calculated incorrectly)

```{r tp_calc_test_1, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(h$tp>1.841 && h$tp<4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

Tests that it calculates tp correctly for fish id 136

```{r tp_calc_test_2, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(round(h$tp[h$id==136], digits=3)==3.166)
{
  print("passed")
} else
{
  print("failed")
}
```
  
## Tests for fcl_calc

Test to make sure that FCL was calculated correctly for site "SCR"

```{r fcl_calc_test_1, echo=T}
tpdf <- na.omit(tp_calc(fish, 3.4))
h <- fcl_calc(tpdf)
if(round(h$fcl[h$site=="SCR"], digits=3)==4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

## Descriptive test for plot_web  

Plot should have delta13C values on the x-axis and delta15N values on the y-axis. delta13C values should be a range of negative numbers, while delta15N values should be a range of positive numbers. There should be separate groups plotted, in the colors blue, red, green, and black.  

```{r plot_web_test_1, echo=T}
tpdf <- tp_calc(fish, 3.4)
h <- plot_web(tpdf)
```
