---
title: "Final Project Tests"
author: "Emily Arsenault"
date: "May 3, 2018"
output: pdf_document
---

## Chunk for importing the functions to test

```{r, echo=T}
source("base_sig.R")
source("tp_calc.R")
source("fcl_calc.R")
source("plot_web.R")
```

## Tests for base_sig

Test that it can find the correct primary producer signature mean for site "WIN"

```{r base_sig_test_1, echo=T}
sitetest <- "WIN"
frac <- 3.4
h <- base_sig(muss, frac)
if(round(h[sitetest,2], digits = 3)==7.287)
{
  print("passed")
} else
{
  print("failed")
}
```

## Tests for tp_calc

Test to make sure that no fish trophic position is less than the minimum trophic position that we observed or greater than the maximum trophic position observed (if it is, it was calculated incorrectly)

```{r tp_calc_test_1, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(h$tp>1.841 && h$tp<4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

Tests that it calculates tp correctly for fish id 136

```{r tp_calc_test_2, echo=T}
fish <- read.csv("fish.csv")
h <- tp_calc(fish, frac=3.4)
if(round(h$tp[h$id==136], digits=3)==3.166)
{
  print("passed")
} else
{
  print("failed")
}
```
  
## Tests for fcl_calc

Test to make sure that FCL was calculated correctly for site "SCR"

```{r fcl_calc_test_1, echo=T}
tpdf <- na.omit(tp_calc(fish, 3.4))
h <- fcl_calc(tpdf)
if(round(h$fcl[h$site=="SCR"], digits=3)==4.463)
{
  print("passed")
} else
{
  print("failed")
}
```

## Descriptive test for plot_web  

In all cases, plot should have delta13C values on the x-axis and delta15N values on the y-axis. 

In all cases, delta13C values should be a range of negative numbers, while delta15N values should be a range of positive numbers. 

In all cases, there should be separate groups plotted; however, the number of groups and the color scheme may vary with the dataset. Color scheme may be changed directly in function specs. Grouping variable (dataframe column) may be changed using the "group" argument when executing the function.

```{r plot_web_test_1, echo=T}
tpdf <- tp_calc(fish, 3.4)
h <- plot_web(tpdf, tpdf$feeding)
```
