---
title: Effect of hydrogeomorphology on trophic complexity of the Upper Mississippi River (USA)
author: E. R. Arsenault, J. H. Thorp, M. D. Delong, and R. E. Bowes
date: 03 May 2018
output: 
  pdf_document:
    fig_caption: yes
fontsize: 12pt
bibliography: datacarpentry.bib
csl: ecology.csl
---

[Note: I have simplified my project a bit to make it more feasible for the timeframe. While the original sampling design mentioned in my proposal included sampling sub-habitats within each of 4 major sites (MMR, SCR, PEP, and WIN), I am ignoring those and just focusing on differences among the 4 sites themselves.]

<!--Chunk for installing and loading necessary packages-->
```{r checkpoint, echo=F, include=F, warning=F}

#Dependencies: need internet connection to run checkpoint
#Note: .checkpoint folder should have been added to my GitHub repo, so I hope it runs with less trouble this time!

#installing packages (if not already installed) and loading libraries
projpacks <-  c("checkpoint", "plyr", "mapdata", "RColorBrewer", "grDevices", "maps", "maptools", "ggplot2", "SIBER") #packages used in this project
projinst <- projpacks %in% installed.packages() #finding necessary packages within packages already installed

#installing packages not already installed
if(length(projpacks[!projinst]) > 0) 
{
  install.packages(projpacks[!projinst])
}

#loading all packages (require)
lapply(projpacks, require, character.only = TRUE)

#creating a .checkpoint folder in directory (unless it already exists)
if (!dir.exists("./.checkpoint/"))
{
  dir.create("./.checkpoint/")
}
date <- "2018-04-01"
checkpoint(date,checkpointLocation = "./")
checkpoint(date)

#Outputs: loaded packages for all functions in file, checkpoint for package reproducibility
```

<!--Chunk for sourcing external scripts-->
```{r functions, echo=F, include=F, warning=FALSE}

#Dependencies: external scripts, each with 1 function

source("base_sig.R")
source("tp_calc.R")
source("fcl_calc.R")
source("gen_col.R")
source("plot_web.R")
source("layman_calc.R")

#Output: available functions base_sig(), tp_calc(), fcl_calc(), gen_col(), plot_web(), and layman_calc()
```

<!--Chunk for importing the data-->
```{r import, echo=F}

#Dependencies: csv files "muss.csv" (primary consumer data) and "fish.csv" (fish food web data)

muss <- read.csv("muss.csv", header = T) #primary producer signature data (mussels)
fish <- read.csv("fish.csv", header = T) #fish consumer richness, length, mass, and isotope data

mussdf <- base_sig(muss = muss, frac = 3.4)
tpdf <- tp_calc(mussdf, fish = fish, frac = 3.4)
fcldf <- fcl_calc(tpdf)
laydf <- layman_calc(tpdf)

#Output: muss and fish dataframes (CSV files) and mussdf tpdf, fcldf, and laydf dataframes (outputs from functions), data needed for all analyses in this file
```

## **Abstract**

Historically, most research in river ecology has focused on investigating processes in great detail over small spatial scales. More recently, studies have begun to suggest the importance of large-scale studies spanning whole river networks in understanding patterns ans processes in lotic ecosystems. In particular, viewing rivers as patchy, discontinuous segments divided by unique hydrogeomorphological characteristics provides a unique perspective on analyzing habitat complexity and its effect on biotic interactions like food webs. Here, we examined trophic relationships among fish at sites of varying hydrogeomorphic complexity along the Upper Mississippi River (UMR). The highest hydrogeomorphic complexity was observed at a naturally anastomosing section of the UMR. However, not all metrics of food web complexity were highest at that site. Results show that, while total niche space was greatest within the anastomosing section of the UMR, trophic and food source diversity was highest further downstream within the main channel. The results of this study suggest that hydrogeomorphic complexity may contribute to some aspects of food chain complexity in rivers, but more studies are needed to understand the specific drivers of food web complexity in this system.

## **Introduction**

Recent developments in ecology suggest that the understanding of ecological processes may benefit from an expanded focus at the macrosystem scale, especially in the context of global climate change and species invasions [@Williams2013; @Grabowski2016]. Past studies in river ecology have focused primarily on microhabitats and relatively small stream reaches, guided by conceptual models often over-emphasizing the longitudinal gradient of rivers as a predictor of food web interactions (e.g., River Continuum Concept, [@Vannote1980]). In more recent attempts to increase spatial scale, researchers have employed meta-analyses and extrapolation techniques; however, these types of studies may not be adequate in themselves for understanding natural systems [@Hilderbrand2004].  

The Riverine Ecosystem Synthesis (RES), proposed by @Thorp2006, offered a new framework for understanding river ecosystems by “portray[ing] rivers as downstream arrays of large hydrogeomorphic patches (e.g., constricted, braided and floodplain channel areas) formed by catchment geomorphology and climate.” This model integrated concepts from the models before it, but refuted the River Continuum Concept [@Vannote1980] by suggesting a discontinuous nature of river networks in contrast to the predictable and continuous gradient originally proposed. The theoretical foundation of RES was based on concepts from hierarchical patch dynamics [@Wu1995; @Townsend1989; @Poole2002], hydrogeomorphology [@Thoms2002], and the multiple dimensionality of rivers [@Ward1989].

Recently, the RES concepts of hydrogeomorphological patches (FPZs) and the hierarchical nature of a landscape have been directly tied to ecological hypotheses and studies of the river macrosystem [@Thorp2014; @McCluney2014]. An ecological macrosystem can be defined as “a hierarchically organized, integrated terrestrial, inland aquatic and/or marine ecological unit of large spatial extent (c. $10^{2}$-$10^{6}$ $km^{2}$ or more depending on the types and sizes of ecosystems present) whose temporal interactions within the unit and with regional through global processes are especially significant over periods of decades to millennia” [@Thorp2014]. This concept can be applied to many different systems, but in river ecology, the river macrosystem can be defined as a network of interacting riverscape and floodscape patches. Riverscapes include all currently active main channels, side channels, and backwaters of rivers, while floodscapes consist of the normally dry floodplains, floodplain lakes, wetlands, and oxbow lakes [@Thorp2006].  

Here, we conduct a large-scale food web study to better understand the effects of hydrogeomorphology on ecosystem function within a macrosystem framework. Specifically, we investigated the effect of river hydrogeomorphic complexity on fish food web complexity in four unique FPZs of the Upper Mississippi River to determine whether or not there was an effect of habitat complexity on fish food web complexity. We predicted that food chain length, as well as food web complexity, as quantified using six different Layman's metrics, would increase with site habitat complexity.

## **Methods**

### *Study Sites*

This study was conducted between the months of May and July 2010 in the Upper Mississippi River (UMR) watershed. Sites were located along a large geographic gradient and represented a variety of hydrogeomorphic characteristics. Four major areas along the UMR were sampled; upstream to downstream, we sampled a tributary (St. Croix River), a reservoir (Lake Pepin), a naturally anastomosing area (near Winona, MN), and a main channel site (above the confluence of the UMR and the Illinois River; Figure 1).

<!--Chunk for making a map of the study sites-->
```{r map, echo=F, fig.height=8, fig.width=4, warning=F, fig.cap="Study sites located along four different hydrogeomorphic zones of the Mississippi River"}

#Dependencies: csv file with lat and long coordinates for each site

latuppmiss <- c(45.5675, 44.50, 43.8442, 39.4408)
lonuppmiss <- c(-92.8647, -92.30, -91.2923, -91.0259)
sites <- c("SCR", "PEP", "WIN", "MMR")

uppmissdf <- as.data.frame(c(latuppmiss, lonuppmiss, sites))
uppmiss_map <- map(database = "state", region = c("minnesota", "wisconsin", "missouri", "iowa", "illinois"), 
                   col=("gray90"), fill=TRUE, mar = c(0, 0, 1, 0))
points(lonuppmiss, latuppmiss, pch=16, col='black', cex=1)
with(uppmissdf, text(lonuppmiss-.55, latuppmiss, cex=0.7, labels=sites))
box()
map.scale(ratio = FALSE, metric = FALSE)

#Output: map of Upper Mississippi River study sites (uppmiss_map)
```

### *Quantification of Site Complexity*

Hydrogeomorphic measurements were taken at transects placed at 500 m intervals along a 10 km reach located within the sampling area for each of the four UMR sites. We measured the wetted width, or the linear distance between banks at current water level, at each transect. We also determined the number of channels and islands that were intersected by each transect, as well as the sinuosity of each reach. A combined sum of these 4 measurements were used to quantify channel complexity. Ranks for these variables were then summed to provide a single measure of hydrogeomorphic complexity for each site.

### *Fish and Mussel Collection*

Suspension-feeding mussels were hand-collected in shallow areas of each site. A representative fish community sample was collected from each 10-km site reach using boat electrofishing. Fishes and mussels were identified to species and classified into functional feeding groups based on regional fisheries manuals [@Etnier1993; @Pflieger1997]. We released all native mussels after the sampling a small amount of protenaceous periostracum, but invasive zebra mussels were collected whole and sacraficed. Small fish were collected whole, but larger fish were released after a small muscle plug was collected non-lethally from the dorsal tissue. 

### *Stable Isotope Analysis*

Tissues from all collected fishes and mussels were used for bulk-tissue stable isotope analysis of carbon and nitrogen ($\delta^{13}C$ and $\delta^{15}N$). All collected tissues were rinsed, dried in an oven for 48 hr at 60&deg;C , and homogenized into a fine powder using a Wig-L-Bug &reg; Mixer-Amalgamator. Prepared samples were held in a dessicator until submitted to the W. M. Keck Paleoenvironmental and Environmental Stable Isotope Laboratory (University of Kansas, Lawrence, KS, USA) for bulk-tissue stable isotope analysis of carbon and nitrogen. Isotope data for each sample included  $\delta^{13}C$ and $\delta^{15}N$ signatures, which were determined by the relative difference in isotope ratio between each sample and a known standard (Vienna Pee Dee Belemnite for C or atmospheric nitrogen for N) using the formula: $$\delta X=((Rsample-Rstandard)-1)*1000$$ where X is either $^{13}C$ or $^{15}N$ and the corresponding isotopic ratio (R) is either $^{13}C/^{12}C$ or $^{15}N/^{14}N$. All isotopic signature data is represented using per mille (&permil;) notation. 

### *Statistical Analysis*

All statistical analyses were conducted in R version 3.4.1 [@R2014]. Consumer (fish) trophic position was calculated by the formula: $$Trophic Position = [(\delta^{15}N consumer - \delta^{15}N herbivore)/3.4]+1]$$ where mean primary producer signatures were calculated using herbivore (mussel) signatures collected from each Mississippi River site. Food chain length was defined as the maximum trophic position for each site. In addition, we calculated six Layman's metrics using the SIBER package in R [@SIBER]: $\delta^{15}n$ range, $\delta^{13}C$ range, total area of the convex hull for each site, mean distance to centroid, mean nearest neighbor distance (NND), and the standard deviation of the NND, as detailed in @Layman2007. The fish food web of the Upper Mississippi River was represented using an isotope biplot of $\delta^{13}C$ and $\delta^{15}N$ signature values. 

## **Results**

[Note: I have simplified this project in a way that it will not really have any statistical meaning right now, but I hope to have set this up in a way that I should be able to just fill in additional site data and re-knit the Rmd file later to update.]

### *Site Characterization*

Average hydrogeomorphic complexity varied among UMR sites. From upstream to downstream, the average hydrogeomorphic complexity scores for each site were: 2.53 for SCR (headwaters), 4.43 for PEP (reservoir), 6.01 for WIN (anastomosing area), and 3.51 for MMR (main channel). 

In total, `r nrow(na.omit(tpdf))` fish representing `r length(levels(tpdf$species))` were collected from `r length(levels(tpdf$site))` `r length(levels(tpdf$site))` sites (`r levels(tpdf$site)`) for bulk-tissue stable isotope analysis (see Supplemental Materials for more details). At the headwaters of the Mississippi River at site SCR, we collected `r length(unique(fish$species[fish$site=="SCR"]))` species. At the artificial reservoir, site PEP, `r length(unique(fish$species[fish$site=="PEP"]))` fish species were collected. Further downstream, at site WIN, a naturally anastomosing section of the UMR, `r length(unique(fish$species[fish$site=="WIN"]))` fish species were collected. Finally, at MMR, the site furthest downstream along the main channel of the UMR, we collected `r length(unique(fish$species[fish$site=="MMR"]))` different species.

### *Food Web Structure*

Mean trophic position for all sites along the Upper Mississippi was `r round(mean(tpdf$tp), digits = 2)`. Food chain length ranged from `r round(min(fcldf$fcl), digits=2)` at `r fcldf$site[fcldf$fcl==min(fcldf$fcl)]` to `r round(max(fcldf$fcl), digits=2)` at `r fcldf$site[fcldf$fcl==max(fcldf$fcl)]`.  

<!--Chunk for generating table of Layman's metrics-->
```{r layman, echo=F}

#Dependencies: layman_calc() function, data generated from tp_calc function

laytab <- data.frame(round(laydf$MMR$metrics, digits = 2), round(laydf$PEP$metrics, digits = 2), round(laydf$SCR$metrics, digits= 2), round(laydf$WIN$metrics, digits = 2))
colnames(laytab) <- c("MMR", "PEP", "SCR", "WIN")
knitr::kable(laytab, caption = "Layman's metrics for four sites along the Upper Mississippi River (SCR, PEP, WIN, and MMR). Layman's metrics include d15N range (dY_range), d13C range (dX_range), total area (TA), centroid distance (CD), mean nearest neighbor distance (NND), and standard deviation of the nearest neighbor distance (SDNND)")

#Output: data frame with 6 Layman's metrics: N range, C range, total area, centroid distance, mean nearest neighbor distance, and standard deviation of the nearest neighbor distances
```


<!--Chunk for making isotope plot by site-->
```{r siteplot, echo=F, results="hide", fig.cap="Biplot of d13C and d15N bulk tissue isotope signatures for fish collected in four FPZs of the Mississippi River (n=514). Points are colored according to site FPZ, with MMR in red, PEP in orange, SCR in green, and WIN in blue"}

#Dependencies: plot_web() function, data frame generated by tp_calc function

plot_web(tpdf)

#Output: plot with x=d13C and y=d15N for fish collected in four different zones of the Mississippi River
```

<!--Chunk for correlation plot of hydrogeomorphic complexity and food web complexity-->
```{r complexity, echo=F, include=F}

#Dependencies: hydrogeomorphic complexity measurements (given, not calculated), outputs from fcl_calc() and layman_calc()

hgmcomplex <- c(3.51, 4.43, 2.53, 6.01) #in order of: MMR, PEP, SCR, WIN

#To add in later with additional site data:
#summary(lm([METRIC]~hgmcomplex))
#plot([METRIC]~hgmcomplex)

#Output: (Will be) regression plots and statistics, but does not make sense to include right now with just 4 data points
```

Results from six Layman's metrics suggest differences in food web structure among sites of different hydrogeomorphology (Table 1; Figure 2). Maximum N range (`r round(laydf$MMR$metrics["dY_range"], digits = 2)`) and maximum C range (`r round(laydf$MMR$metrics["dX_range"], digits = 2)`) both occurred at site MMR. The largest total area (`r round(laydf$WIN$metrics["TA"], digits = 2)`) and the maximum centroid distance among the sites (`r round(laydf$WIN$metrics["CD"], digits = 2)`) were both observed at site WIN. Maximum mean nearest neighbor distance occurred at site PEP (`r round(laydf$PEP$metrics["NND"], digits = 2)`), while the highest standard deviation of the nearest neighbor distance (`r round(laydf$MMR$metrics["SDNND"], digits = 2)`) was observed at site MMR (Table 1).  

## **Discussion**

Recent expansions to river conceptual models suggest that rivers can be more accurately viewed as patches of varying hydrogeomorphic complexity, where river segments with related hydrogeomorphologic characteristics are expected to be more similar than river segments related by their geographic proximity alone [@Thorp2006]. Here, our results suggest that hydrogeomorphic complexity may be a possible predictor of certain, but not all, aspects of food web structure.

Hydrogeomorphic complexity values varied among UMR sites. As expected, the greatest hydrogeomorphic complexity was observed at site WIN, a naturally anastomosing site with many channels and islands. The second highest complexity occurred at site PEP; however, the complexity value at this site may have been inflated due to its very high wetted width value, since PEP is an artificial reservoir (Lake Pepin). MMR, the site furthest downstream on the main channel of the UMR had the third highest complexity value, and SCR, a site in the headwaters of the UMR, had the lowest complexity score overall. These values were also as expected. Both MMR and SCR were largely single-channel areas with few islands and side channels, but the MMR site downstream had a larger wetted width value than the smaller river segments in the St. Croix headwaters, which raises its complexity score.

Food chain length was highest at SCR, the site with the lowest hydrogeomorphic complexity, which does not support our hypothesis that food chain length would increase with habitat complexity. The individual that was observed to be at the top of the food web at SCR was a smallmouth bass (*Micropterus dolomieu*), which is a species known for its aggressive predation [@MacRae2001]. It is possible that food chain length might have been higher than expected in these headwater areas because that section of the UMR is partially protected by Wild River State Park. Higher forest cover and natural habitat in the surrounding riparian areas might allow large fish more opportunities to feed on terrestrial subsidies to the river, such as small mammals. The shortest food chain length was observed at Lake Pepin (PEP). While lakes tend to have longer food chains than rivers [@VanderZanden2007], the artificial nature of the reservoir may have altered this natural trend. More work should be conducted to better understand how anthropogenic disturbances may be disrupting the natural aquatic food web in this area.

Different aspects of food web complexity, as quantified using Layman's metrics [@Layman2007], suggested different patterns when compared to site complexity scores. MMR had the widest spread of both $\delta^{13}C$ and $\delta^{15}N$ values, which suggests a food web supported by a large diversity of different food sources, which supports many different trophic roles, from top predators down to detritivores. Total area (TA) is a measure of total niche space available at a site [@Layman2007]. Site WIN had the highest TA and CD, which may suggest that hydrogeomorphic complexity does support complex ecological roles in the environment, even if those differences in trophic roles do not ultimately support the highest food chain length along the UMR. The fact that WIN also exhibited the highest species richness supports this finding. WIN also showed the lowest trophic redundancy, as quantified using mean nearest neighbor distance (NND), which suggests that a larger niche space allowed for many unique (non-redundant) trophic roles in the system. The standard deviation of the NND, which is a proxy for evenness of species packing in a community, was lowest in WIN and highest in MMR. This suggests that the MMR food web is tightly clustered within isotopic space, while the WIN food web has links that display a wide and uneven spread of taxa (Figure 2). This may suggest that trophic roles have become more specialized in the hydrogeomorphically complex site, WIN, while fish employ generalist strategies in the main channel site MMR.

In conclusion, our results suggest that hydrogeomorphic complexity may have a positive effect on trophic complexity. WIN, the most hydrogeomorphically complex UMR site, had the highest species richness and also showed high trophic complexity based on Layman's metrics. We recommend more studies to determine whether or not these same patterns can be observed in other river macrosystems.

## **Author Contributions**

J. H. T. and M. D. D. formulated the question and collected all field data. R. E. B. prepared and processed the samples. E. R. A. conducted statistical analyses and drafted the manuscript.

## **Supplemental Materials**

See supplement for abundance information for each species of mussel and fish collected along all sites of the Upper Mississippi River.

## **References**