---
title: Effect of hydrogeomorphology on trophic complexity of the Upper Mississippi River (USA)
author: E. R. Arsenault, J. H. Thorp, M. D. Delong, and R. E. Bowes
date: 03 May 2018
output: pdf_document
fontsize: 12pt
bibliography: datacarpentry.bib
csl: ecology.csl
---

Note: I have simplified the research question a bit since my project proposal for the sake of clarity & my ability to actually focus on the programming aspect of the assignment.

[JUST MAJOR ZONES, NO MEGAHABITAT STUFF]

<!--Chunk for installing and loading necessary packages-->
```{r checkpoint, echo=F, include=F, warning=F}

#Dependencies: need internet connection to run checkpoint
#Note: .checkpoint folder should have been added to my GitHub repo, so I hope it runs with less trouble this time!

#installing packages (if not already installed) and loading libraries
projpacks <-  c("checkpoint", "plyr", "mapdata", "RColorBrewer", "maps", "maptools", "ggplot2") #packages used in this project
projinst <- projpacks %in% installed.packages() #finding necessary packages within packages already installed

#installing packages not already installed
if(length(projpacks[!projinst]) > 0) 
{
  install.packages(projpacks[!projinst])
}

#loading all packages (require)
lapply(projpacks, require, character.only = TRUE)

#creating a .checkpoint folder in directory (unless it already exists)
if (!dir.exists("./.checkpoint/"))
{
  dir.create("./.checkpoint/")
}
date <- "2018-04-01"
checkpoint(date,checkpointLocation = "./")
checkpoint(date)

#Outputs: loaded packages for all functions in file
```

<!--Chunk for importing the data-->
```{r import, echo=F}

#Dependencies: csv files "muss.csv" (primary consumer data) and "fish.csv" (fish food web data)

muss <- read.csv("muss.csv", header = T) #primary producer signature data (mussels)
fish <- read.csv("fish.csv", header = T) #fish consumer richness, length, mass, and isotope data

#Output: muss and fish dataframes, data needed for all analyses in this file
```

## **Abstract**

Historically, many studies in river ecology have focused on investigating processes in great detail over small spatial scales. More recently, studies have begun to suggest the importance of large-scale studies in river ecology. In particular, viewing rivers through their unique hydrogeomorpholgies provides a unique perspective on analyzing habitat complexity and its effect on biotic interactions like food webs. Here, we examined trophic relationships among fish at sites of varying hydrogeomorphic complexity along the Upper Mississippi River. [MAJOR FINDING 1] [MAJOR FINDING 2] The results of this study suggest that hydrogeomorphic complexity is a driver of food chain complexity in rivers, a finding that has important implications for river conservation and management as we strive to promote complexity in this important river ecosystem.

## **Introduction**

Recent developments in ecology suggest that the understanding of ecological processes may benefit from an expanded focus at the macrosystem scale, especially in the context of global climate change and species invasions [@Williams2013; @Grabowski2016]. Past studies in river ecology have focused primarily on microhabitats and relatively small stream reaches, guided by conceptual models often over-emphasizing the longitudinal gradient of rivers as a predictor of food web interactions (e.g., River Continuum Concept, [@Vannote1980]). In more recent attempts to increase spatial scale, researchers have employed meta-analyses and extrapolation techniques; however, these types of studies may not be adequate in themselves for understanding natural systems [@Hilderbrand2004].  

The Riverine Ecosystem Synthesis (RES), proposed by @Thorp2006, offered a new framework for understanding river ecosystems by “portray[ing] rivers as downstream arrays of large hydrogeomorphic patches (e.g., constricted, braided and floodplain channel areas) formed by catchment geomorphology and climate.” This model integrated concepts from the models before it, but refuted the River Continuum Concept [@Vannote1980] by suggesting a discontinuous nature of river networks in contrast to the predictable and continuous gradient originally proposed. The theoretical foundation of RES was based on concepts from hierarchical patch dynamics [@Wu1995; @Townsend1989; @Poole2002], hydrogeomorphology [@Thoms2002], and the multiple dimensionality of rivers [@Ward1989].

Recently, the RES concepts of hydrogeomorphological patches (FPZs) and the hierarchical nature of a landscape have been directly tied to ecological hypotheses and studies of the river macrosystem [@Thorp2014; @McCluney2014]. An ecological macrosystem can be defined as “a hierarchically organized, integrated terrestrial, inland aquatic and/or marine ecological unit of large spatial extent (c. $10^{2}$-$10^{6}$ $km^{2}$ or more depending on the types and sizes of ecosystems present) whose temporal interactions within the unit and with regional through global processes are especially significant over periods of decades to millennia” [@Thorp2014]. This concept can be applied to many different systems, but in river ecology, the river macrosystem can be defined as a network of interacting riverscape and floodscape patches. Riverscapes include all currently active main channels, side channels, and backwaters of rivers, while floodscapes consist of the normally dry floodplains, floodplain lakes, wetlands, and oxbow lakes [@Thorp2006].  

Here, we conduct a large-scale study of food web dynamics to investigate the effects of hydrogeomorphology on ecosystem function within a macrosystem framework. In particular, we investigated the effect of river hydrogeomorphic complexity on fish food web complexity in four unique FPZs of the Upper Mississippi River to determine whether or not there was an FPZ effect on fish food web complexity in the Upper Mississippi River. We predicted that fish food web complexity, as measured by several common metrics, would increase with increasing habitat complexity within an FPZ.

## **Methods**

### *Study Sites*

This study was conducted between the months of May and July 2010 in the Upper Mississippi River (UMR) watershed. Sites were located along a large geographic gradient and represented a variety of hydrogeomorphic characteristics. Four major areas along the UMR were sampled; upstream to downstream, we sampled a tributary (St. Croix River), a reservoir (Lake Pepin), a naturally anastomosing area (near Winona, MN), and a main channel site (above the confluence of the UMR and the Illinois River; Figure 1). We also sampled multiple megahabitat types within each of these major zones: main channels, side channels, and closed backwaters.

<!--Chunk for making a map of the study sites-->
```{r map, echo=F, fig.height=8, fig.width=4, warning=F, fig.cap="Study sites located along four different hydrogeomorphic zones of the Mississippi River"}

#Dependencies: csv file with lat and long coordinates for each site

#placeholder map, which I will continue to work on for the final project (mapping in R is NOT intuitive!)
latuppmiss <- c(45.5675, 44.50, 43.8442, 39.4408)
lonuppmiss <- c(-92.8647, -92.30, -91.2923, -91.0259)
sites <- c("SCR", "PEP", "WIN", "MMR")
uppmissdf <- as.data.frame(c(latuppmiss, lonuppmiss, sites))
uppmiss_map <- map(database = "state", region = c("minnesota", "wisconsin", "missouri", "iowa", "illinois"), col=('gray90'), fill=TRUE, mar = c(0, 0, 1, 0))
points(lonuppmiss, latuppmiss, pch=16, col='black', cex=1)
with(uppmissdf, text(lonuppmiss-.55, latuppmiss, cex=0.7, labels=sites))
box()
map.scale(ratio = FALSE, metric = FALSE)

#Output: map of Upper Mississippi River study sites (uppmiss_map)
```

### *Quantification of Site Complexity*

Hydrogeomorphic measurements were taken at transects placed at 500-m intervals along a 10-km reach located within each sampling reach. The wetted width, or the linear distance between banks at current water level, was measured at each transect. We also determined the number of channels and islands intersected by each transect, as well as the sinuosity of the 10-km sections. A combined sum of these 4 measurements were used these measurements to quantify channel complexity. Ranks for these variables were then summed to provide a single measure of hydrogeomorphic complexity for each site.

### *Fish and Mussel Collection*

Suspension-feeding mussels were hand-collected in shallow areas of each site. A representative fish community sample was collected from each site along a 10-km reach using boat electrofishing. Fishes and mussels were identified to species and classified into functional feeding groups based on regional fisheries manuals [@Etnier1993; @Pflieger1997]. For all native mussels, non-lethal collection of protenaceous periostracum was conducted, but invasive zebra mussels were collected whole and sacraficed. Small fish were collected whole, but larger fish were released after a small muscle plug was collected non-lethally from the dorsal tissue. 

### *Stable Isotope Analysis*

Tissues from all collected fishes and mussels were used for bulk-tissue stable isotope analysis of carbon and nitrogen ($\delta^{13}C$ and $\delta^{15}N$). All collected tissues were rinsed, dried in an oven for 48 hr at 60&deg;C , and homogenized into a fine powder using a Wig-L-Bug &reg; Mixer-Amalgamator. Prepared samples were held in a dessicator until submitted to the W. M. Keck Paleoenvironmental and Environmental Stable Isotope Laboratory (University of Kansas, Lawrence, KS) for bulk tissue stable isotope analysis. Isotope data for each sample included  $\delta^{13}C$ and $\delta^{15}N$ signatures, which were determined by the relative difference in isotope ratio between each sample and a known standard (Vienna Pee Dee Belemnite for C or atmospheric nitrogen for N) using the formula: $$\delta X=((Rsample-Rstandard)-1)*1000$$

### *Statistical Analysis*

All statistical analyses were conducted in R [@R2014]. Consumer trophic position was calculated by the formula: $$Trophic Position = [(\delta^{15}N consumer - \delta^{15}N herbivore)/3.4]+1]$$, where mean primary producer signatures were calculated using primary consumer (mussel) signatures collected from each Mississippi River site. Food chain length was defined as the maximum trophic position for each site. The fish food web of the Upper Mississippi River was represented using an isotope biplot of $\delta^{13}C$ and $\delta^{15}N$ signature values. 

## **Results**

<!--Chunk for sourcing external scripts-->
```{r, echo=F}

#Dependencies: external scripts, each with 1 function

source("base_sig.R")
source("tp_calc.R")
source("fcl_calc.R")
source("plot_web.R")
source("layman_calc.R")

#Output: available functions base_sig(), tp_calc(), fcl_calc(), plot_web(), and layman_calc()
```

### *Quantification of Hydrogeomorphic Complexity*

The average ranks from cumulative variables were 7.32 (Upper Mississippi at LaCrosse), 4.70 (Upper Miss at Winona), 4.43 (Lake Pepin), 4.07 (lower site in our most downstream hydrogeomorphic zone of the Mississippi), 2.94 (upper site in the most downstream hydrogeomorphic zone), 2.86 (most downstream site within the St. Croix hydrogeomorphic zone), and 2.20 (upstream site within the St. Croix zone). The complexity value for Lake Pepin site was high because of width and wetted width. 

### *Organism Collections*

In total, `r nrow(na.omit(tpdf))` fish were collected from `r levels(tpdf$site)` for bulk tissue stable isotope analysis. Average trophic position for all sites along the Upper Mississippi was `r round(mean(tpdf$tp), digits = 2)`. Food chain length ranged from `r round(min(fcldf$fcl), digits=2)` at `r fcldf$site[fcldf$fcl==min(fcldf$fcl)]` to `r round(max(fcldf$fcl), digits=2)` at `r fcldf$site[fcldf$fcl==max(fcldf$fcl)]`. 

### *Analysis of Food Web Complexity*

An isotope biplot suggests differences in food web ecology among sites (Figure 2).

<!--Chunk for making isotope plot by site-->
```{r siteplot, echo=F, results="hide", fig.cap="Biplot of d13C and d15N bulk tissue isotope signatures for fish collected in four FPZs of the Mississippi River (n=514). Points are colored according to site FPZ, with MMR in red, PEP in blue, SCR in green, and WIN in black."}

#Dependencies: plot_web function, data frame generated by tp_calc function

plot_web(tpdf, tpdf$site)

#Output: plot with x=d13C and y=d15N for fish collected in four different zones of the Mississippi River
```

<!--Chunk for making isotope plot by megahabitat-->
```{r habitatplot, echo=F, results="hide", fig.cap="Biplot of d13C and d15N bulk tissue isotope signatures for fish collected in four FPZs of the Mississippi River (n=514). Points are colored according to site FPZ, with MMR in red, PEP in blue, SCR in green, and WIN in black."}

#Dependencies: plot_web function, data frame generated by tp_calc function

plot_web(tpdf, tpdf$habitat)

#Output: plot with x=d13C and y=d15N for fish collected in four different zones of the Mississippi River
```

<!--Chunk for making isotope plot by feeding group-->
```{r feedingplot, echo=F, results="hide", fig.cap="Biplot of d13C and d15N bulk tissue isotope signatures for fish collected in four FPZs of the Mississippi River (n=514). Points are colored according to site FPZ, with MMR in red, PEP in blue, SCR in green, and WIN in black."}

#Dependencies: plot_web function, data frame generated by tp_calc function

plot_web(tpdf, tpdf$feeding)

#Output: plot with x=d13C and y=d15N for fish collected in four different zones of the Mississippi River
```

<!--Chunk for generating table of Layman's metrics-->
```{r layman, echo=F}

#Dependencies:

layman_calc(tpdf, tpdf$site)

#Output:
```

## **Discussion**

Another plug for studying rivers at a large scale

Hydrogeomorphic complexity

Food web complexity relationship with hydrogeomorphic complexity?

## **Author Contributions**

J. H. T. and M. D. D. formulated the question and collected all field data. R. E. B. prepared and processed the samples. E. R. A. conducted statistical analyses and drafted the manuscript.

## **References**